Sparse Function-space Representation \\ of Neural Networks

### 2023-05-24 paper/main.tex
updated section A2

\section{Method details}
\label{app:method}
In this section, we first show in \cref{app:losses} how the dual parameters used for \our are computed for the different losses employed in our experiments, and then we discuss the computational complexity of our method in \cref{app:comp_compl}.
%
\subsection{Computation of the dual parameters for different losses}\todo{just a placeholder}
\label{app:losses}
As shown in \cref{eq:dual_param} in the main paper, our method rely on computation of the dual parameters. We show here how we can derive them in general for the family of exponential likelihoods. By rewriting the loss in the natural form, we have,
\begin{equation}
 \ell(h(f), y) = -\log p(y \mid f) =  A(f)  - \langle t(y), f \rangle
\end{equation}
where $A(f)$ is the log partition function, $h(\cdot)$ is the inverse link function, $t(y)$ are the sufficient statistics, and $f$ denotes the neural network output before the link function. 
%given the loss functions we use $t(y) = y$. 
Then, taking the first and second derivatives of the loss in this form become:
\begin{equation}\label{eq:dual_param_app}
  -\nabla_{f} \log p(y_i \mid f) |_{f=f_i} =  A'(f_i) - y_i
  \quad \text{and} \quad
  \nabla^2_{f\!f} - \log p(y_i \mid f) |_{f=f_i} = A''(f_i).
\end{equation}
% The second derivative of the log partition function is easily calculated and
In this section, we provide the derivatives needed for the losses used in this work, \ie, the mean squared error (MSE) likelihood, used for the regression tasks, the Bernoulli likelihood, for binary classification, and the categorical likelihood, used for multi-class classification.

For regression, the negative log likelihood comes from the MSE, and for a single input it can be written as:
\begin{equation}
	\ell(h(f), y) = \frac{1}{2} (y_i -f_i)^2.% -\log \Norm( h(f) | y, \sigma^2) 	= - \log 
\end{equation}
where the inverse link function is simply $h(f) = f$. Thus, the first and second derivatives are
\begin{equation}
	\nabla_f \ell(\sigma(f), y_i)|_{f=f_i} =  f_i - y_i \quad \text{ and } \quad  \nabla_{f\!f} \ell(h(f), y_i)|_{f=f_i} = 1
\end{equation}

In binary classification, we deal with Bernoulli likelihoods. The inverse link function is the sigmoid function $h(f) = \sigma(f) = \frac{1}{1 + e^{-f}}$, and we can then write the derivates as follows
%\begin{equation}
%\begin{aligned}
%	\ell(h(f), y) 
%	&= -\log \left( \sigma(f)^y (1 - \sigma(f))^{1-y} \right) \\
%	&= - y \log(\sigma(f)) - (1-y) \log(1- \sigma(f)) \\
%	&= - y f + \log(1 + e^f) 
%\end{aligned}
%\end{equation}
%Then, it follows that the the derivative we are interested in are
\begin{equation}
	\nabla_f \ell(h(f), y_i)|_{f=f_i} = \sigma(f_i) - y_i  \quad \text{ and } \quad  \nabla_{f\!f} \ell(h(f), y_i)|_{f=f_i} = \sigma(f_i) (1 - \sigma(f_i)).
\end{equation}

Moving to the multi class classification, the function outputs is no more a scalar but a vect ogits $f_\weights(\vx_i) = \bm{f _i} \in \R^C$. The inverse link function for the categorical loss is the softmax function $\textrm{Softmax}(\bm{f}) : \R^C \mapsto \R^C$ defined as $h(\bm{f}) = \textrm{Softmax}(\bm{f}) = \frac{\exp(\bm{f})}{ \sum^C_{c=1} \exp(\bm{f}_c)} \in \R^C$. If we write the targets as a one-hot  encoded vector $\vy_i$, we can write the negative log likelihood as:
\begin{equation}
\begin{aligned}
	\ell(h(\bm{f}),\vy_i) &= - \log\left( \vy_i^\T \textrm{Softmax}(\bm{f})\right)\\
	% &= - y_i \cdot \log \left( \left[ \textrm{Softmax}(\bm{f}) \right]_i \right)
\end{aligned}
\end{equation}
% where $\left[\textrm{Softmax}(\bm{f})\right]_j = \frac{\exp(\bm{f}_j)}{\sum_{c=1}^{C} \exp(\bm{f}_c)} \in \R$ defines the $j$\textsuperscript{th} entry of the logits fed through the $\textrm{Softmax}(\cdot)$ function. 
Finally, we can write the first and second derivatives as follows
\begin{equation}
\begin{aligned}
	[\nabla_{\bm{f}} \ell(h(\bm{f}), \vy_i)|_{\bm{f}=\bm{f}_i}]_j &= \left[\textrm{Softmax}(\bm{f}) - \vy_i  \right]_j \\
	\left[\textrm{diag}\left(\nabla_{\bm{f}\!\bm{f}} \ell(h(\bm{f}), \vy_i)|_{\bm{f}=\bm{f}_i}\right)\right]_{j} &= \left[\textrm{Softmax}(\bm{f})\right]_j (1 - \left[\textrm{Softmax}(\bm{f})\right]_j),
\end{aligned}
\end{equation}

where $\left[\textrm{Softmax}(\bm{f})\right]_i = \frac{\exp(\bm{f}_i)}{\sum_{j=1}^{C} \exp(\bm{f}_j)} \in [0, 1]$ defines the $i$\textsuperscript{th} entry of the logits fed through the $\textrm{Softmax}(\cdot)$ function. Finally, we can write the first and second derivatives as follows

\begin{equation}
\begin{aligned}
	\nabla_f \ell(h(f), y_i)|_{f=f_i} &= - y_i + \left[ \textrm{Softmax}(\bm{f}) \right]_i \\
	\nabla_{f\!f} \ell(h(f), y_i)|_{f=f_i} &= \left[\textrm{Softmax}(\bm{f})\right]_i (1 - \left[\textrm{Softmax}(\bm{f})\right]_i).
\end{aligned}
\end{equation}
where we only need the $\textrm{diag}$ given our independence assumption.

@@

\subsection{Computational complexity}
\label{app:comp_compl}

% Big O
% train neural network
% compute the NTK -> from the NTK paper
% we outer product of the kzi O(NM^2)
% inversion O(M^3)

% Practical implementation
% mention that we do that in batches not full N 
% role of jitter 
% role of clamping
% GPU A100 Nvidia

@@

\subsection{Extending the CL regularizer to multi-class settings} 
\label{sec:cl_multioutput}
%
In a multi-class classification setting, the neural network can be seen as a function approximator that maps inputs $\vx \in \inputDomain$ to the logits $\textit{\textbf{f}} \in \outputDomain$, \ie, $f_\mathbf{w}: \inputDomain \to \outputDomain$. As a consequence, the Jacobian matrix is $\Jac{\weights}{\vx} \coloneqq \left[ \nabla_\weights f_\weights(\vx)\right]^\top \in \R^{C \times P}$ and the kernel of the associated sparse GP derived in \cref{sec:sfr}, $\kappa(\vx, \vx')$ is a tensor in $\R^{C \times P \times P}$. This means that we need to compute the dual parameters and the matrix $\bar{\MB}^{-1}_{t, c}$ for all the function's outputs. If we denote $\kappa_c(\vx, \vx') \in \R$, as the kernel related to the $c$\textsuperscript{th} output function, \ie, the logit for class $c$, we can write the dual parameters as
%
\begin{equation} \textstyle
  \valpha_{\vu, c}  =  \sum_{i=1}^N  \vkzic \, \hat{\alpha}_{i, c} \in \R^{M}
  \quad \text{and} \quad
  \MBeta_{\vu, c} =  \sum_{i=1}^N \vkzic \,\hat{\beta}_{i, c} \, \vkzic^{\T} \in \R^{M \times M} ,    
\label{eq:dual_sparse}
\end{equation}
%
where $\vkzic$ is the kernel $\kappa_c(\cdot, \cdot)$ computed for the inducing points $\vz \in \mathcal{Z}$ and the input point $\vx_i$ for class $c$. In this setting, the dual variables are
\begin{equation}
\begin{aligned}
  \hat{\alpha}_{i,c} &\coloneqq  \left[\hat{\valpha}_{i}\right]_c = \left[\nabla_{f}\log p(y_i \mid f) |_{f=f_i} \right]_c \in \R,\\
  \hat{\beta}_{i,c} &\coloneqq \left[\textrm{diag}(\hat{\vbeta}_i)\right]_c = \left[ - \nabla^2_{ff}\log p(y_i \mid f) |_{f=f_i} \right]_c \in \R,
\end{aligned}
\end{equation}
where the operator $[ \cdot ]_c$ is taking the $c$\textsuperscript{th} element of a vector. In fact, in the multi-class setting $\hat{\valpha}_{i}$ is a vector in $\R^C$ and $\hat{\vbeta}_{i}$ a matrix in $\R^{C \times C}$.
For the \our-regularizer, at the end of each task, we compute the matrix $\bar{\MB}^{-1}_{t, c}$ as 
\begin{equation}\textstyle
 	\bar{\MB}^{-1}_{t, c} = \MKzzc^{-1} \MBeta_{\vu, c} \MKzzc^{-1} \in \R^{M \times M}, 
 	\quad \text{s.t.} \quad
 	\MBeta_{\vu, c} =  \sum_{i \in \dataset_t} \vkzi \,\hat{\beta}_{i, c} \, \vkzi^{\T}.
\end{equation}
The kernel computation is based on the relationship between the observed data and the parametrized function. For this reason, computing the kernel $\kappa_c$ for classes $c$, for which the model has not observed any data yet, is not meaningful.  \todo{probably better to word this differently}
In practice, for our method we define $\bar{\MB}^{-1}_{t, c}$ to be $\bar{\MB}^{-1}_{t, c} = \MI_M$ $\forall c \not\in \mathcal{C}_t$, where $\mathcal{C}_t$ is the set of classes observed up to task t. For example, for S-MNIST, after task 1 the set of observed classes for which we will compute the kernel is $\mathcal{C}_1 = \{0, 1\}$, then after task 2 it becomes $\mathcal{C}_1 = \{0, 1, 2, 3\}$, et cetera. 
Finally, we can rewrite $\mathcal{R}_\mathrm{\our}$ as follows
\begin{equation}\label{eq-cl-regularizer}%\textstyle
  \mathcal{R}_\mathrm{\our}(\weights, \mathcal{M}) = \frac{1}{2} \sum_{s=1}^{t-1} \sum_{c=1}^{C} \frac{1}{M} 
	\left\lVert 
	\vu_c - \vu_{s,c} % f_{\weights_{s}}(\MZ_s)
	\right\rVert_{\bar{\MB}^{-1}_{s}},
\end{equation}
%
with $\vu_c \in \R^M$ and $u_{i, c} = \left[ f_{\weights}(\vz_{i}) \right]_c, \forall {\vz_i} \in \mathcal{Z}_s$.

@@

The results reported in \cref{tbl:cl_table_1}, are obtained using the following hyperparameters and repeating the run five times with different seeds. The results are in-line with what has been reported in previous work, notably even better for some of the baseline methods. Thus, we consider the presented results to give a realistic picture of the relative performance of the different methods, and a demonstrating a clear case for applicability of the \our approach in continual learning.

>>>>

\section{Method details}
\label{app:method}
In this section, we first show in \cref{app:losses} how the dual parameters used for \our are computed for the different losses employed in our experiments, and then we discuss the computational complexity of our method in \cref{app:comp_compl}.
%
\subsection{Computation of the dual parameters for different losses}\todo{just a placeholder}
\label{app:losses}
As shown in \cref{eq:dual_param} in the main paper, our method rely on computation of the dual parameters. We show here how we can derive them in general for the family of exponential likelihoods. By rewriting the loss in the natural form, we have,
\begin{equation}
 \ell(h(f), y) = -\log p(y \mid f) =  A(f)  - \langle t(y), f \rangle
\end{equation}
where $A(f)$ is the log partition function, $h(\cdot)$ is the inverse link function, $t(y)$ are the sufficient statistics, and $f$ denotes the neural network output before the link function. 
%given the loss functions we use $t(y) = y$. 
Then, taking the first and second derivatives of the loss in this form become:
\begin{equation}\label{eq:dual_param_app}
  \nabla_{f} - \log p(y_i \mid f) |_{f=f_i} =  - A'(f_i) y_i
  \quad \text{and} \quad
  \nabla^2_{f\!f} - \log p(y_i \mid f) |_{f=f_i} = A''(f_i).
\end{equation}
\todo{check the signs with the main paper}
% The second derivative of the log partition function is easily calculated and
In this section, we provide the derivatives needed for the losses used in this work, \ie, the mean squared error (MSE) likelihood, used for the regression tasks, the Bernoulli likelihood, for binary classification, and the categorical likelihood, used for multi-class classification.

For regression, the negative log likelihood comes from the MSE, and for a single input it can be written as:
\begin{equation}
	\ell(h(f), y) = \frac{1}{2} (y -f).% -\log \Norm( h(f) | y, \sigma^2) 	= - \log 
\end{equation}
where the inverse link function is simply $h(f) = f$. Thus, the first and second derivatives are
\begin{equation}
	\nabla_f \ell(\sigma(f), y_i)|_{f=f_i} = y_i - f_i \quad \text{ and } \quad  \nabla_{f\!f} \ell(h(f), y_i)|_{f=f_i} = 1
\end{equation}

In binary classification, we deal with Bernoulli likelihoods. The inverse link function is the sigmoid function $h(f) = \sigma(f) = \frac{1}{1 + e^{-f}}$, and we can then write the negative log likelihood as
\begin{equation}
\begin{aligned}
	\ell(h(f), y) 
	&= -\log \left( \sigma(f)^y (1 - \sigma(f))^{1-y} \right) \\
	&= - y \log(\sigma(f)) - (1-y) \log(1- \sigma(f)) \\
	&= - y f + \log(1 + e^f) 
\end{aligned}
\end{equation}
Then, it follows that the the derivative we are interested in are
\begin{equation}
	\nabla_f \ell(h(f), y_i)|_{f=f_i} = - y_i + \sigma(f_i) \quad \text{ and } \quad  \nabla_{f\!f} \ell(h(f), y_i)|_{f=f_i} = \sigma(f_i) (1 - \sigma(f_i)).
\end{equation}

Moving to the multi class classification, the function outputs is no more a scalar but a vector of logits $\bm{f} \in \R^C$. The inverse link function for the categorical loss is the softmax function $\textrm{Softmax}(\bm{f}) : \R^C \mapsto \R^C$ defined as $h(\bm{f}) = \textrm{Softmax}(\bm{f}) = \frac{\exp(\bm{f})}{ \sum^C_{j=1} \exp(\bm{f}_j)} \in \R^C$. If we write the targets $y_i$ as a one-hot $C$-shaped one-hot encoding vector, we can write the negative log likelihood as:
\begin{equation}
\begin{aligned}
	\ell(h(\bm{f}),\vy) &= - \log\left( \vy^\T \textrm{Softmax}(\bm{f})\right)\\
	&= - y_i \cdot \log \left( \left[ \textrm{Softmax}(\bm{f}) \right]_i \right)
\end{aligned}
\end{equation}
where $\left[\textrm{Softmax}(\bm{f})\right]_i = \frac{\exp(\bm{f}_i)}{\sum_{j=1}^{C} \exp(\bm{f}_j)} \in [0, 1]$ defines the $i$\textsuperscript{th} entry of the logits fed through the $\textrm{Softmax}(\cdot)$ function. Finally, we can write the first and second derivatives as follows

\begin{equation}
\begin{aligned}
	\nabla_f \ell(h(f), y_i)|_{f=f_i} &= - y_i + \left[ \textrm{Softmax}(\bm{f}) \right]_i \\
	\nabla_{f\!f} \ell(h(f), y_i)|_{f=f_i} &= \left[\textrm{Softmax}(\bm{f})\right]_i (1 - \left[\textrm{Softmax}(\bm{f})\right]_i).
\end{aligned}
\end{equation}

@@

\subsection{Computational complexity}
\label{app:comp_compl}
% Big O
% train neural network
% compute the NTK -> from the NTK paper
% we outer product of the kzi O(NM^2)
% inversion O(M^3)
In this section, we analyze the computational complexity of \our. We first need to train a neural network $f_\weights$ and then compute the posterior covariance matrix that characterizes our sparse GP representation of the model. The first step then requires the usual complexity cost for training a neural network, where defining the cost of a forward pass (and backward pass
%\todo{these two should be similar, we can't cite the NTK paper were they are using this notation unfortunately}
) as $[F\!P]$, this can roughly sketched as $\mathcal{O}(2  E  N[F\!P])$, where $E$ is the number of training epochs and  $N$ the number of training samples. Thus, there is no additional cost involved in the training phase of the neural network.

Given the trained network, we construct the covariance function  which is defined through the NTK~\citep{jacot2018neural}. We relied on the Jacobian contraction implementation using PyTorch~\citep{paszke2019pytorch}. To compute the kernel Gram matrix for a batch of $B$ samples with this implementation has a computational complexity of $\mathcal{O}( B C [F\!P] + B^2 C^2 P)$, where $C$ is the number of NN outputs and $P$ is the number of parameters of the neural network. In \cref{eq:dual_sparse}, we compute $\MB_\vu$ iterating through all the $N$ training points to compute the $\vkzi$ and then use them to compute the outer product, that has a cost of $\mathcal{O}(M^2)$, with $M$ being the number of inducing points following the notation used in the main paper. The total cost of this operation becomes $\mathcal{O}(N M^2)$, which is also the dominating cost of our approach.

% Practical implementation
% mention that we do that in batches not full N 
% role of jitter 
% role of clamping
% GPU A100 Nvidia
In practical terms, some care needs to be taken during the implementation, both in terms of memory consumption and numerical stability. For the former, we batched over the number of training samples, $N$, which keeps the memory consumption in check without need for any further approximations. The computation of the terms in \cref{eq:dual_sparse_post} require the inversion of an $M \times M$ matrix, that we implemented in a numerically stable manner by using Cholesky factorization. We also introduce a small jitter and clip of the GGN matrix for numerical stability where needed (see discussions in \cref{app:experiments}). 
 We run our experiments GPU clusters equipped with NVIDIA A100 GPUs.

@@

\subsection{Extending the CL regularizer to multi-class settings} 
\label{sec:cl_multioutput}
%
In a multi-class classification setting, the neural network can be seen as a function approximator that maps inputs $\vx \in \inputDomain$ to the logits $\textit{\textbf{f}} \in \outputDomain$, \ie, $f_\mathbf{w}: \inputDomain \to \outputDomain$. As a consequence, the Jacobian matrix is $\Jac{\weights}{\vx} \coloneqq \left[ \nabla_\weights f_\weights(\vx)\right]^\top \in \R^{C \times P}$ and the kernel of the associated sparse GP derived in \cref{sec:sfr}, $\kappa(\vx, \vx')$ is a tensor in $\R^{C \times P \times P}$. This means that we need to compute the dual parameters and the matrix $\bar{\MB}^{-1}_{t, c}$ for all the function's outputs. If we denote $\kappa_c(\vx, \vx') \in \R$, as the kernel related to the $c$\textsuperscript{th} output function, \ie, the logit for class $c$, we can write the dual parameters as
%
\begin{equation} \textstyle
  \valpha_{\vu, c}  =  \sum_{i=1}^N  \vkzic \, \hat{\alpha}_{i, c} \in \R^{M}
  \quad \text{and} \quad
  \MBeta_{\vu, c} =  \sum_{i=1}^N \vkzic \,\hat{\beta}_{i, c} \, \vkzic^{\T} \in \R^{M \times M} ,    
\label{eq:app_dual_sparse}
\end{equation}
%
where $\vkzic$ is the kernel $\kappa_c(\cdot, \cdot)$ computed for the inducing points $\vz \in \mathcal{Z}$ and the input point $\vx_i$ for class $c$. In this setting, the dual variables are
\begin{equation}
\begin{aligned}
  \hat{\alpha}_{i,c} &\coloneqq  \left[\hat{\valpha}_{i}\right]_c = \left[\nabla_{f}\log p(y_i \mid f) |_{f=f_i} \right]_c \in \R,\\
  \hat{\beta}_{i,c} &\coloneqq \left[\textrm{diag}(\hat{\vbeta}_i)\right]_c = \left[ - \nabla^2_{ff}\log p(y_i \mid f) |_{f=f_i} \right]_c \in \R,
\end{aligned}
\end{equation}
where the operator $[ \cdot ]_c$ is taking the $c$\textsuperscript{th} element of a vector. In fact, in the multi-class setting $\hat{\valpha}_{i}$ is a vector in $\R^C$ and $\hat{\vbeta}_{i}$ a matrix in $\R^{C \times C}$.
For the \our-regularizer, at the end of each task, we compute the matrix $\bar{\MB}^{-1}_{t, c}$ as 
\begin{equation}\textstyle
 	\bar{\MB}^{-1}_{t, c} = \MKzzc^{-1} \MBeta_{\vu, c} \MKzzc^{-1} \in \R^{M \times M}, 
 	\quad \text{s.t.} \quad
 	\MBeta_{\vu, c} =  \sum_{i \in \dataset_t} \vkzi \,\hat{\beta}_{i, c} \, \vkzi^{\T}.
\end{equation}
The kernel computation is based on the relationship between the observed data and the parametrized function. For this reason, computing the kernel $\kappa_c$ for classes $c$, for which the model has not observed any data yet, is not meaningful.  \todo{probably better to word this differently}
In practice, for our method we define $\bar{\MB}^{-1}_{t, c}$ to be $\bar{\MB}^{-1}_{t, c} = \MI_M$ $\forall c \not\in \mathcal{C}_t$, where $\mathcal{C}_t$ is the set of classes observed up to task t. For example, for S-MNIST, after task 1 the set of observed classes for which we will compute the kernel is $\mathcal{C}_1 = \{0, 1\}$, then after task 2 it becomes $\mathcal{C}_1 = \{0, 1, 2, 3\}$, et cetera. 
Finally, we can rewrite $\mathcal{R}_\mathrm{\our}$ as follows
\begin{equation}\label{eq-cl-regularizer}%\textstyle
  \mathcal{R}_\mathrm{\our}(\weights, \mathcal{M}) = \frac{1}{2} \sum_{s=1}^{t-1} \sum_{c=1}^{C} \frac{1}{M} 
	\left\lVert 
	\vu_c - \vu_{s,c} % f_{\weights_{s}}(\MZ_s)
	\right\rVert_{\bar{\MB}^{-1}_{s}},
\end{equation}
%
with $\vu_c \in \R^M$ and $u_{i, c} = \left[ f_{\weights}(\vz_{i}) \right]_c, \forall {\vz_i} \in \mathcal{Z}_s$.

@@

The results reported in \cref{tbl:cl_table_1}, are obtained using the following hyperparameters and repeating the runs five times with different seeds. The results are in-line with what has been reported in previous work, notably even better for some of the baseline methods. Thus, we consider the presented results to give a realistic picture of the relative performance of the different methods, and a demonstrating a clear case for applicability of the \our approach in continual learning.
